---
title: "Processing the datafiles"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Currently, HaDeX is limited to the `cluster` file from DynamX 3.0 or 2.0 and `tables` file from HDeXaminer. Each data file type needs a different way of processing, as explained below.

## Processing data from DynamX

Let's investigate how the data processing of the `cluster` file from DynamX is conducted on the example file from the package.

### Initial processing of the data 

```{r}
library(HaDeX)
dat <- read_hdx(system.file(package = "HaDeX", "HaDeX/data/KD_180110_CD160_HVEM.csv"))
head(dat, 10)
```

As you can see, the data file has a very specific structure and is not informative yet. In the file we have `m/z` values for each `z` value (charge) for each time point for each state of each peptide, repeated as many times as measurement was repeated (each measurement should be repeated at least 3 times). The process of measurement and the basics of HDX experiment is described in `Basics of the HDX-MS` article.

Our aim is to have one result with uncertainty of the measurement per each peptide in each biological state in each measured time point - data in this format allows further calculations e.q. calculating deuterium uptake values.

Within each replicate of the measurement (we recognize each replicate by the `File` value) the `m/z` values are provided for each possible `z` value. The `m/z` values are in the `Center` column, as it is a centroid calculated from the spectrum. Firstly, we have to calculate mass value, measured experimentally:

$$expMass = z \times (Center-protonMass)\tag{1}$$
The value is reduced by ... because ...

After this operation there are as many results as there are replicates of the measurement. To aggregate it, let's calcultate the mean weighted by intensity (because ...) .

$$aggMass = \frac{1}{N}\sum_{k = 1}^{N}Inten_k{\cdot}pepMass_k\tag{2}$$.

Where:

* $aggMass$ - mass of the peptide [Da], average from replicates,
* $Inten_k$ - intensity of the measurement,
* $pepMass$ - mass of the peptide, calculated using Equation 1.

As we use the aggregated result from the replicates, we need to calculate an uncertainty associated with the measurement. We use the mean value as the final result, so we need to calculate error as a standard deviation of the mean, according to the Equation 3:

$$u(\vec{x}) = \sqrt{\frac{ \sum_{i=1}^n \left( x_{i} - \overline{x} \right)^2}{n(n-1)}}\tag{3}$$.

Where:

* $x_{i}$ - measured value from a replicate,
* $\overline{x}$ - mean value from all of the replicates,
* $n$ - number of replicates.

Now we have the format we want for further calculations.

### Calculating the deuterium uptake

In HaDeX package, we are showing the calculated values in different forms. All of them are provided with associated uncertainty of the measurement. All of the uncertainties are derived from the formula - the Law of propagation of uncertainty:

$$u_{c}(y) = \sqrt{\sum_{k} \left[ \frac{\partial y}{\partial x_{k}} u(x_{k}) \right]^2}$$

#### Deuterium uptake

Deuterium uptake is the increase of the mass of the peptide in time $t$. The minimal exchange control $m_{t_0}$ is mass measured directly after adding the buffer, and $m_t$ is the mass measured in chosen time point $t$. The value is in Daltons [Da].

$$D = m_{t} - m_{t_0}$$
The uncertainty associated with deuterium uptake [Da]:

$$u_c(D) = \sqrt{u(m_t)^2 + u(m_{t_0})^2}$$

#### Fractional deuterium uptake

Fractional deutetium uptake is the ratio of the increase of the mass in time $t$ to the maximal possible increase. The maximal exchange control $m_{t_{100}} is measured after long time period (chosen by experimentator, usually 1440 h), and is assumed to be measured after the exchange stops. The minimal exchange control $m_{t_0}$ is mass measured directly after adding the buffer, and $m_t$ is the mass measured in chosen time point $t$. This is a percentage value [%]. 

$$D_{frac} = \frac{m_{t} - m_{t_0}}{m_{t_{100}} - m_{t_0}}$$

The uncertainty associated with fractional deuterium uptake [%]:

$$u_{c}(D_{frac}) = \sqrt{ \left[ 
\frac{1}{m_{t_{100}}-m_{t_0}} u(m_{t}) \right]^2 + 
\left[ \frac{m_{t} - m_{t_{100}}}{(m_{t_{100}}-m_{t_0})^2} u(m_{t_0}) \right]^2 + 
\left[ \frac{m_{t_0} - m_{t}}{(m_{t_{100}}-m_{t_0})^2} u (m_{t_{100}}) \right]^2}$$

#### Theoretical deuterium uptake

Theoretical deuterium uptake is the increase of mass in time $t$ compared with theoretical mass value of the peptide ($MHP$ -	mass of the singly charged monoisotopic molecular ion) and $m_t$ is the mass measured in chosen time point $t$. This value is in daltons [Da]:

$$D_{theo} = m_{t} - MHP$$

The uncertainty associated with theoretical deuterium uptake [Da] (as the $MHP$ value is a constans, without any uncertainty):

$$u(D_{theo}) = u(m_t)$$

#### Theoretical fractional deuterium uptake

Theoretical fractional deuterium uptake is the ratio of the increase of mass in time $t$ compared with theoretical mass value of the peptide to the theoretical possible increase of the mass, based on the maximal possible uptake of the peptide. This value is a percentage [%].

$$D_{theo, frac} = \frac{m_{t}-MHP}{MaxUptake \times protonMass}$$

The uncertainty associated with theoretical fractional deuterium uptake [%]:

$$u(D_{theo, frac}) = \left|\frac{1}{MaxUptake \times protonMass} u(D_{t}) \right|$$

#### Differential values

Differential value is the way to see how the deuterium mass increase differs between two states. It also allows to see if the difference is statistically important (we will talk about it later). This value is calculates as the difference between previously calculated (in chosen form) deuterium uptake of first state and second one. 

$$diff = D_{1} - D_{2}$$
The uncertainty associated with the difference of deuterium uptake:

$$u_{c}(diff) = \sqrt{u(D_{1})^2 + u(D_{2})^2}$$

The convenient way to present results calculated as described above is comparison plot and differential plot (Woods' plot).

### Additional information 

If the file contains modified peptides - the value from column `Modification` is added to the value from column `State` and is treated as a new biological condition. The further aspects of the analysis are the same as for non-modified peptides.

## Processing data from HDeXaminer
TO BE WRITTEN