---
title: "HaDeX package"
author: "Weronika Pucha&#322;a, Micha&#322; Burdukiewicz"
date: "15.02.2019"
output: 
  rmarkdown::html_vignette:
    toc: true
    header-includes:
      -\usepackage{amsmath}
vignette: >
  %\VignetteIndexEntry{HaDeX package - an overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo = FALSE, message = FALSE, results='asis'}
library(HaDeX)
library(ggplot2)
library(knitr)
library(DT)
opts_chunk$set(fig.width = 7, fig.height = 5)
dat <- read_hdx(system.file(package = "HaDeX", "HaDeX/data/KD_190304_Nucb2_EDTA_CaCl2_test02_clusterdata.csv"))
# source(system.file(package = "HaDeX", "HaDeX/data-work.R"))
```

# Hydrogen/Deuterium Exchange Mass Spectrometry Data 

Hydrogen-deuterium mass spectrometry (HDX-MS) is a staple tool for monitoring dynamics and interactions of proteins. Due to the sheer size of the HDX-MS results, the data analysis require a dedicated software suite. However, the majority of existing tools provides only point-and-click interfaces to black-box models or does not offer a complete workflow. We propose \textbf{HaDeX}, a novel tool for processing, analysis and visualisation of HDX-MS experiments. \textbf{HaDeX} covers the whole analytic process, including preliminary data exploration, quality control and generation of publication-quality figures. The reproducibility of the whole procedure is ensured with advanced reporting functions.

# General data manipulation

## Input file format 

The \textbf{HaDeX} web server relies only on a single data format: DynamiX datafile. An user can customize each step of the analysis with his own input parameters.

## Transformation of data and measurement uncertainty

Uploaded datafile should fulfills following requirements in order for package function to behave correctly:

```{r warning=FALSE, message=FALSE, echo = FALSE}

datatable(
  data = data.frame(Name = c("Protein", "Start", "End", "Sequence", "Modification", "Fragment", "MaxUptake", "MHP", "State", "Exposure", "File", "z", "RT", "Inten", "Center"),
                    Type = c("Character", "Integer", "Integer", "Character", "Logic", "Logic", "Numeric", "Numeric", "Character", "Numeric", "Character", "Integer", "Numeric", "Numeric", "Numeric")),
  rownames = FALSE,
  options = list(dom = "t", autoWidth = TRUE))

```

In upcoming version of HaDeX more datafile types will be supported.

### Protein coverage

First information user can get is protein sequence reconstruction based on supplemented data. Although the sequence should be known to the experimenter, it is uselful to see how the sequence was covered in experiment. Function reconstruct_sequence() does that:

```{r waring=FALSE}

reconstruct_sequence(dat)

```

Unknown amino is marked as x.

### Data preparation for plots

$$expMass = z*(Center-protonMass)\tag{1}$$
where:

- $expMass$ - experimentally obtained mass, calculated

- $z$ - charge

- $Center$ - data from experiment

Data from file containing repeated measurement is aggregated into one result using formula: 

$$expMass = \frac{Inten*expMass}{N}\tag{2}$$
where:

- $Inten$ - Intensivity

- $N$ - number of repetitions

In input file is data from more than one measurement. The mean value for each time in each state is treated as a result. This means that the uncertainty is calculated using the standard deviation of the mean:

$$u(x) = \sqrt{\frac{ \sum_{i=1}^n \left( x_{i} - \overline{x} \right)^2}{n}}\tag{3}$$
where:

- $x_{i}$ - measurement
 
- $\overline{x}$ - mean value

- $n$ - number of measurements

Experimental deuteration level (function of three variables):

$$D = \frac{D_{t} - D_{0}}{D_{100} - D_{0}}\tag{4}$$

where:

- $D$ - deuteration level

- $D_{0}$ - deuteration in in time (0 or close to 0, measured) - mean value obtained as explained before

- $D_{100}$ - deuteration in out time (possibly big, at least 1440) - mean value obtainted as explained before

- $D_{t}$ - deuteration measured in chosen time - mean value obtainted as explained before


Law of propagation of uncertainty.

$$u_{c}(y) = \sqrt{\sum_{k} \left[ \frac{\partial y}{\partial x_{k}} u(x_{k}) \right]^2}\tag{5}$$


Uncorrelatedness is assumed.

Translated with equation 1:

$$u_{c}(D) = \sqrt{\sum_{k} \left[ \frac{\partial D}{\partial D_{k}} u(D_{k}) \right]^2 }\tag{6}$$

where:

- $k \in {0, t, 100}$

- $D_{k}$ - deuteration in k time as standard deviation of the mean value 

- $u(D_{k})$ - uncertainty associated with $D_{k}$

Then, expanding the equation:

$$u_{c}(D) = \sqrt{ \left[ \frac{1}{D_{100}-D_{0}} u(D_{t}) \right]^2 + \left[ \frac{D_{t} - D_{100}}{(D_{100}-D_{0})^2} u(D_{0}) \right]^2 + \left[ \frac{D_{0} - D_{t}}{(D_{100}-D_{0})^2} u(D_{100}) \right]^2}\tag{7}$$

For theoretical calculation (function of one variable), deuteration:

$$D = \frac{D_{t}-MHP}{MaxUptake*protonMass}\tag{8}$$

where:

- $D_{t}$ - deuteration measured in chosen time

- $MHP$ - theoretical mass for peptide (constant)

- $MaxUptake$ - max proton uptake for peptide (constant)

- $protonMass$ - mass of a proton (constant)

and the uncertainty for function of one variable looks like:

$$u(y) = \left| \frac{dy}{dx} u(x) \right|\tag{9}$$

and in our case:

$$u(D) = \left|\frac{1}{MaxUptake*protonMass} u(d_{t}) \right|\tag{10}$$

For differential plot in Woods format, deueration level in one state $(D_{2})$ is subtracted from deuteration level in other state $(D_{1})$:

$$diff = D_{1} - D_{2}\tag{11}$$

and the uncertainty of function of two variables (based on previous equation):

$$u_{c}(diff) = \sqrt{u(D_{1})^2 + u(D_{2})^2}\tag{12}$$
All of the previous calculations are done by function prepare_dataset(). Example of use:

```{r warning=FALSE}
calc_dat <- prepare_dataset(dat,
                            in_state_first = "gg_Nucb2_EDTA_0.001",
                            chosen_state_first = "gg_Nucb2_EDTA_25",
                            out_state_first = "gg_Nucb2_EDTA_1440",
                            in_state_second = "gg_Nucb2_CaCl2_0.001",
                            chosen_state_second = "gg_Nucb2_CaCl2_25",
                            out_state_second = "gg_Nucb2_CaCl2_1440") 
```

### Comparison plot

The levels of deuteration along with uncertainty intervals are shown on comparison plot. HaDeX provides both experimental and theroetical levels of deuteration. Example of use:

- theoretical:

```{r warning=FALSE}
comparison_plot(calc_dat = calc_dat,
                theoretical = TRUE,
                state_first = "Nucb2 Factor 1",
                state_second = "Nucb2 Factor 2") +
  labs(title = "Theoretical fraction exchanged in \nstate comparison in 25 min time")
```

- experimental:

```{r warning=FALSE}
comparison_plot(calc_dat = calc_dat,
                theoretical = FALSE,
                state_first = "Nucb2 Factor 1",
                state_second = "Nucb2 Factor 2") +
  labs(title = "Fraction exchanged in state comparison \nin 25 min time")
```

### Woods plot

Woods plot format shows diference between results of two different states as described by equation 11. HaDeX provides both experimental i theoretical calculations. Example of use:

- theoretical:

```{r warning=FALSE}
woods_plot(calc_dat = calc_dat,
           theoretical = TRUE) +
  labs(title = "Theoretical fraction exchanged between \nstates in 25 min time")
```

- experimental:

```{r warning=FALSE }
woods_plot(calc_dat = calc_dat,
           theoretical = FALSE) +
  labs(title = "Theoretical fraction exchanged between \nstates in 25 min time")
```

## Additional tools

HaDeX provides additional tools for assement of experiments. 

### Quality control 

Function quality_control() shows how the mean uncertainty per peptide changes with time points of an experiment. Example of use:

```{r warning=FALSE}
result <- quality_control(dat = dat,
                          state_first = "gg_Nucb2_EDTA",
                          state_second = "gg_Nucb2_CaCl2", 
                          chosen_time = 25, 
                          in_time = 0.001)

```

As the function returns data.frame with calculated value, here is an example of result visualization:

```{r warning=FALSE}
ggplot(result[result["time"]>=1,]) + 
  geom_line(aes(x = time, y = avg_err_state_first, color = "Average error (first state)")) +
  geom_line(aes(x = time, y = avg_err_state_second, color = "Average error (second state)")) +
  scale_x_log10() +
  labs(x = "log(time) [min]", y = "Average uncertainty", title = "Uncertainty change") + 
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom",
        legend.title = element_blank())
```

# Example workflow

1. File import
2. Sequence reconstruction
3. Woods plots.

## Example workflow 1

```{r}
library(HaDeX)

# file import
dat_1 <- read_hdx(system.file(package = "HaDeX", "HaDeX/data/KD_180110_CD160_HVEM.csv"))

# protein sequence reconstruction
reconstruct_sequence(dat_1)

# calculate data
calc_dat_1 <- prepare_dataset(dat = dat_1,
                              in_state_first = "CD160_0.001",
                              chosen_state_first = "CD160_1",
                              out_state_first = "CD160_1440",
                              in_state_second = "CD160_HVEM_0.001",
                              chosen_state_second = "CD160_HVEM_1",
                              out_state_second = "CD160_HVEM_1440")                             

# theoretical comparison plot - relative values
comparison_plot(calc_dat = calc_dat_1,
                theoretical = TRUE,
                relative = TRUE, 
                state_first = "CD160",
                state_second = "CD160_HVEM")

# experimental comparison plot - relative values
comparison_plot(calc_dat = calc_dat_1,
                theoretical = FALSE,
                relative = TRUE,
                state_first = "CD160",
                state_second = "CD160_HVEM")

# theoretical comparison plot - absolute values
comparison_plot(calc_dat = calc_dat_1,
                theoretical = TRUE,
                relative = FALSE, 
                state_first = "CD160",
                state_second = "CD160_HVEM")

# experimental comparison plot - absolute values
comparison_plot(calc_dat = calc_dat_1,
                theoretical = FALSE,
                relative = FALSE,
                state_first = "CD160",
                state_second = "CD160_HVEM")
  

# theoretical Woods plot - relative values
woods_plot(calc_dat = calc_dat_1,
           theoretical = TRUE, 
           relative = TRUE) +
  coord_cartesian(ylim = c(-.2, .2))

# experimental Woods plot - relative values
woods_plot(calc_dat = calc_dat_1,
           theoretical = FALSE,
           relative = TRUE) +
  coord_cartesian(ylim = c(-.2, .2))

# theoretical Woods plot - absolute values
woods_plot(calc_dat = calc_dat_1,
           theoretical = TRUE, 
           relative = FALSE) +
  labs(title = "Theoretical fraction exchanged between states in 1 min time") 

# experimental Woods plot - absolute values
woods_plot(calc_dat = calc_dat_1,
           theoretical = FALSE,
           relative = FALSE) +
  labs(title = "Fraction exchanged between states in 1 min time") 

# quality control - relative values 
(result <- quality_control(dat = dat_1,
                state_first = "CD160",
                state_second = "CD160_HVEM", 
                chosen_time = 1, 
                in_time = 0.001))

# example quality control visualisation
library(ggplot2)
ggplot(result) + 
  geom_line(aes(x = time, y = avg_err_state_first, color = "Average error (first state)")) +
  geom_line(aes(x = time, y = avg_err_state_second, color = "Average error (second state)")) +
  scale_x_log10() +
  ylim(0, 0.05) + 
  labs(x = "log(time) [min]", y = "Average uncertainty", title = "Uncertainty change in out time") +
  theme(legend.title = element_blank(),
        legend.position = "bottom")

```

## Example workflow 2

```{r}
library(HaDeX)

# file import
dat_2 <-  read_hdx(system.file(package = "HaDeX", "HaDeX/data/KD_190304_Nucb2_EDTA_CaCl2_test02_clusterdata.csv"))

# protein sequence reconstruction
reconstruct_sequence(dat_2)

# calculate data
calc_dat_2 <- prepare_dataset(dat = dat_2,
                              in_state_first = "gg_Nucb2_EDTA_0.001",
                              chosen_state_first = "gg_Nucb2_EDTA_25",
                              out_state_first = "gg_Nucb2_EDTA_1440",
                              in_state_second = "gg_Nucb2_CaCl2_0.001",
                              chosen_state_second = "gg_Nucb2_CaCl2_25",
                              out_state_second = "gg_Nucb2_CaCl2_1440")                             

# theoretical comparison plot - relative values
comparison_plot(calc_dat = calc_dat_2,
                theoretical = TRUE,
                relative = TRUE,
                state_first = "Nucb2 Factor 1",
                state_second = "Nucb2 Factor 2") +
  labs(title = "Theoretical fraction exchanged in \nstate comparison in 25 min time")

# experimental comparison plot - relative values 
comparison_plot(calc_dat = calc_dat_2,
                theoretical = FALSE,
                relative = TRUE,
                state_first = "Nucb2 Factor 1",
                state_second = "Nucb2 Factor 2") +
  labs(title = "Fraction exchanged in \nstate comparison in 25 min time")

# theoretical comparison plot - absolute values
comparison_plot(calc_dat = calc_dat_2,
                theoretical = TRUE,
                relative = FALSE,
                state_first = "Nucb2 Factor 1",
                state_second = "Nucb2 Factor 2") +
  labs(title = "Theoretical fraction exchanged in \nstate comparison in 25 min time")

# experimental comparison plot - absolute values 
comparison_plot(calc_dat = calc_dat_2,
                theoretical = FALSE,
                relative = FALSE,
                state_first = "Nucb2 Factor 1",
                state_second = "Nucb2 Factor 2") +
  labs(title = "Fraction exchanged in \nstate comparison in 25 min time")

# theoretical Woods plot - relative values
woods_plot(calc_dat = calc_dat_2,
           theoretical = TRUE,
           relative = TRUE) + 
  labs(title = "Theoretical fraction exchanged between states in 25 min time") +
  coord_cartesian(ylim = c(-.5, .7))

# experimental Woods plot - relative values
woods_plot(calc_dat = calc_dat_2,
           theoretical = FALSE,
           relative = TRUE) +
  labs(title = "Fraction exchanged between states in 25 min time") +
  coord_cartesian(ylim = c(-.5, .7))

# theoretical Woods plot - absolute values
woods_plot(calc_dat = calc_dat_2,
           theoretical = TRUE,
           relative = FALSE) + 
  labs(title = "Theoretical fraction exchanged between states in 25 min time") 

# experimental Woods plot - absolute values
woods_plot(calc_dat = calc_dat_2,
           theoretical = FALSE,
           relative = FALSE) +
  labs(title = "Fraction exchanged between states in 25 min time") 

# quality control
(result <- quality_control(dat = dat_2,
                           state_first = "gg_Nucb2_EDTA",
                           state_second = "gg_Nucb2_CaCl2", 
                           chosen_time = 25, 
                           in_time = 0.001))

# example quality control visualisation - relative values
library(ggplot2)
ggplot(result[result["time"]>=1,]) + 
  geom_line(aes(x = time, y = avg_err_state_first, color = "Average error (first state)")) +
  geom_line(aes(x = time, y = avg_err_state_second, color = "Average error (second state)")) +
  scale_x_log10() +
  labs(x = "log(time) [min]", y = "Average uncertainty", title = "Uncertainty change") + 
  theme(legend.position = "bottom",
        legend.title = element_blank())
```
