---
title: "Code usage"
output: 
  rmarkdown::html_vignette:
    toc: true
    number_sections: true
    always_allow_html: yes
bibliography: HDX.bib
vignette: >
  %\VignetteIndexEntry{Code usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r setup, include = FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(dev = "png", dev.args = list(type = "cairo-png"))
```

## Convention

Purpose of groups of functions in the package:

* calculate_* - calculates one specific value. e.q. deuterium uptake values for specific time point.
* create_* - creates a dataset for set of values e.q. multiple time points.
* plot_* - plots a visualization of provided data, accepts calculate_* and create_*  outputs.
* show_* - subsets the dataset to show specific values, accepts calculate_* and create_*  outputs. 

Variables:

* deut_uptake - deuterium uptake [Da]
* frac_deut_uptake - fractional deuterium uptake [%]
* theo_deut_uptake - theoretical deuterium uptake [Da]
* theo_frac_deut_uptake - theoretical fractional deuterium uptake [%]
* diff_deut_uptake - differential deuterium uptake [Da]
* frac_diff_deut_uptake - theoretical differential fractional deuterium uptake [%]
* theo_diff_deut_uptake - differential deuterium uptake [Da]
* theo_frac_diff_deut_uptake - theoretical differential fractional deuterium uptake [%]


## Example of use

```{r message=FALSE, warning=FALSE, echo=FALSE}
library(HaDeX)
library(magrittr)
```

### Read data file 
```{r, include=TRUE}
dat <- read_hdx(system.file(package = "HaDeX", "HaDeX/data/KD_180110_CD160_HVEM.csv"))
```


```{r}
# dat %>%
#   calculate_state_uptake() %>%
#   plot_state_comparison()
```

```{r}
dat %>%
  calculate_peptide_kinetics(protein = "db_CD160",
                             sequence = "INITSSASQEGTRLN",
                             start = 1,
                             end = 15,
                             states = c("CD160", "CD160_HVEM"),
                             time_0 = 0.001, time_100 = 1440) %>%
  plot_uptake_curve(fractional = TRUE)
```

```{r}
dat %>%
  calculate_peptide_kinetics(protein = "db_CD160",
                             sequence = "INITSSASQEGTRLN",
                             start = 1,
                             end = 15,
                             states = c("CD160", "CD160_HVEM"),
                             time_0 = 0.001, time_100 = 1440) %>%
  show_uptake_data(fractional = TRUE)
```

### Different forms of visualization

Create data frame with uptake values for state in all time points:

```{r}
uptake_dat <- create_state_uptake_dataset(dat)
```

Butterfly:

```{r}
plot_butterfly(uptake_dat)
```

Chiclet:

```{r}
plot_chiclet(uptake_dat)
```

