---
title: "hadexReport"
author: ""
date: ""
output:
  html_document:
    theme: united
    code_folding: hide
    highlight: textmate
    toc: true
    toc_float: true
    number_sections: false
    df_print: paged
---

<style type="text/css">
span { 
      display: block;
      max-width: 100%;
      word-wrap: break-word;
    }
</style>

Report generated on `r Sys.time()` using the [HaDeX](http://github.com/michbur/HaDeX) R package.  

Detected input file: `r ifelse(is.null(input[["input_file"]][["name"]]), "none", input[["input_file"]][["name"]])`.  

# Data description

Protein name : `r as.character(unique(dat()[["Protein"]]))`

Protein sequence : <span>`r protein_sequece_colored()`</span>

Length : `r max_range()`

Coverage : `r paste0(round(100-100*str_count(protein_sequence(), 'x')/max_range(), 2), '%')`

<!--- OVERLAP DISTRIBUTION -->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_overlap_dist"]]){
  
  cat("## Overlap Distribution \n")
  
  stateOverlapDist()

}

```

<!--- OVERLAP DISTRIBUTION DATA -->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_overlap_dist_data"]]){

  cat("## Overlap Distribution Data \n")
  
  stateOverlapDist_data()
  
}

```

<!--- OVERLAP GRAPHICALLY -->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_overlap_graph"]]){
    
  cat("## Overlap Graphically \n")
    
  stateOverlap()
  
}


```

<!--- OVERLAP GRAPHICALLY DATA -->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_overlap_graph_data"]]){
    
  cat("## Overlap Graphically Data \n")  
    
  stateOverlap_data()
  
}


```

<!--- COMPARISON PLOT --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_comparison_plot"]]){
    
  cat("## Comparison Plot \n")
  
  comparison_plot_exp()
    
}


```

<!--- COMPARISON PLOT DATA --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_comparison_plot_data"]]){
    
  cat("## Comparison Plot Data \n")
  
  comparison_plot_data_exp()
    
}


```

<!--- THEORETICAL COMPARISON PLOT --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_theo_comparison_plot"]]){
    
  cat("## Theoretical Comparison Plot \n")
  
  comparison_plot_theo()
    
}


```

<!--- THEORETICAL COMPARISON PLOT DATA --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_theo_comparison_plot_data"]]){
    
  cat("## Theoretical Comparison Plot Data \n")
  
  comparison_plot_data_theo()
    
}


```

<!--- WOODS PLOT --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_woods_plot"]]){
    
  cat("## Woods Plot \n")  
  
  differential_plot_exp()
    
}


```

<!--- WOODS PLOT DATA --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_woods_plot_data"]]){
    
  cat("## Woods Plot Data \n")
  
  differential_plot_data_exp()
    
}


```

<!--- THEORETICAL WOODS PLOT --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_theo_woods_plot"]]){
    
  cat("## Theoretical Woods Plot \n")  
  
  differential_plot_theo()
    
}


```

<!--- THEORETICAL WOODS PLOT DATA --->

```{r echo = FALSE, results = "asis", warning = FALSE, message = FALSE}

if (input[["export_theo_woods_plot_data"]]){
    
  cat("## Theoretical Woods Plot Data \n")
  
  differential_plot_data_theo()
    
}
```

# R Session

```{r, results = "markup"}

sessionInfo()

```